Ludwig User Tutorial

### 2022-04-08 docs/tutorial/tutorial.tex

% We're not including the table of contents just at the moment
%\tableofcontents

\newpage
\lstset{style=terminalverbatim}

@@

The following section outlines the configuration and compilation 
procedure. The configuration step is based on the GNU Compiler Collection gcc.

\begin{enumerate}
\item Go to the top level directory \texttt{/ludwig}: \\
\begin{lstlisting}
$ cd ludwig
\end{lstlisting}
\item Copy the (parallel) configuration file \texttt{lunix-mpicc-default.mk}
to top level directory 
and rename to \texttt{config.mk} (both steps exectued below at once): \\
\begin{lstlisting}
$ cp config/unix-mpicc-default.mk ./config.mk
\end{lstlisting}
Note that you may need to specify the correct path to the MPI wrapper compiler
in this file (assumed to be \texttt{mpicc}).
\item Issue \texttt{make} in the top level directory: \\
\begin{lstlisting}
$ make 
\end{lstlisting}
This creates the executable file \texttt{./Ludwig.exe} in \texttt{/src}. 
Note that more than one thread can be used in the compilation process,
which can lead to a considerable speedup  
\begin{lstlisting}
$ make -j 4
\end{lstlisting} 
If you want a serial build, use the conifguation file
\texttt{config/unix-gcc-default.mk} and include the extra step
\begin{lstlisting}
$ make serial
$ make
\end{lstlisting}

@@

In this section the Poiseuille flow of a simple fluid in a cavity, is outlined. 
The example is a step-by-step guide for performing a short run ($10,000$ time-steps) 
with \texttt{Ludwig} code, followed by a short introduction to the post-processing 
and data analysis procedures. Before starting, the code should be compiled in 
parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.
\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory \texttt{/docs/tutorial/test1} 
to your working directory. Note that the input files in this tutorial are modified versions of the 
reference input file \texttt{input.ref} in \texttt{/src} with a limited set of parameters
for each run.
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 4 ./Ludwig.exe input
\end{lstlisting}

\item You should now see statistics being reported on your standard output and 
output files being created with names of the form \texttt{vel-XXXXXXXX.001-001}. 
These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
\item The first step to the post-processing of the data is to compile the \texttt{extract.c} 
routine in \texttt{/util}. This requires the \texttt{Ludwig} code to be compiled in serial 
(see section \ref{configbuild}). Compile with 
\begin{lstlisting}
$ make extract 
\end{lstlisting} 
and copy the executable \texttt{extract} into your working directory.
\item Post-process the raw data by issuing 
\begin{lstlisting}
$ ./extract -a -i meta_file_name data_file_name_stub
\end{lstlisting} 
replacing the filenames with those of the present case (i.e. \texttt{vel.001-001.meta} and e.g. \texttt{vel-00010000}).
The command line options '\texttt{-a}' and '\texttt{-i}' give output in ASCII format and a lattice index 
for easy visualisation in \texttt{gnuplot}.

\item You should now see files named \texttt{vel-XXXXXXXX}.  The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\
\item To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
the spatial $x$-coordinate on the lattice with the command:
\begin{lstlisting}
gnuplot> p `vel-XXXXXXXX' u 1:5
\end{lstlisting} 
By plotting the data in each file, you should be able to see the velocity profile converging to the final profile as shown in Fig. \ref{fig:velocityProfile}.

In the directory \texttt{/docs/tutorial/test1} a reference file of the final 
output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
\end{enumerate}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\linewidth]{velProf.png}
  \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps.}
  \label{fig:velocityProfile}
  \end{center}
\end{figure}

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test2} 
to your working directory. This input file contains the parameters required for this run. 
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item As before, you should now see statistics being reported on your standard output. But 
in this case two sets of output files are being created, one containing the velocity and 
the other compositional order parameter ($\phi$) data at each lattice point in the simulation. 
In this example, these data are in binary format.
\item It should be possible to use the same exectuable \texttt{extract} from the previous 
example for the  post-processing. Alternatively, compile with 
\begin{lstlisting}
make extract
\end{lstlisting} 
and copy the executable \texttt{extract} into your working directory.
Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
(see section \ref{configbuild}).
\item A python script \texttt{extract.py}
can be used for convenient post-processing of multiple raw data files and time steps.
 Edit \texttt{extract.py} and set:
\begin{enumerate}
\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and 
increments of the velocity data
to be processed.
\item \texttt{vel=1} for velocity post-processing, 
\item \texttt{phi=1} for $\phi$ post-processing,
\item The other flags should be set to zero. 
\end{enumerate}
\item Issue the command:
\begin{lstlisting}
$ python extract.py
\end{lstlisting}
\item You should now see files with the extension \texttt{.vtk} corresponding to the velocity 
and $\phi$ data ready to be visualised using ParaView.
\begin{enumerate}
\item In Paraview open the files containing the $\phi$ data.
\item Click on the group of files in the pipeline browser and apply a contour map using 
Filters/Common/Contour or in the tool bar:

@@

For this third example, the system being used is a that of colloidal particles suspended in 
a liquid crystalline fluid.
The example illustrates the post-processing procedure for a colloidal system in addition to 
demonstrating the ways to visualise liquid crystals in ParaView. 
As before, the code should be compiled in parallel and the executable 
\texttt{Ludwig.exe} should be copied into your working directory.

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test3} to your working directory. We see in this input file that 
there are many more parameters 
than in the previous examples. Specifically, we note that \texttt{colloid\_cell\_min} 
must be set if colloids are present and must be at least $2r + \delta$ where $r$ is colloid 
radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.

 \begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\linewidth]{colloidInit.png}
  \caption{A diagrammatic representation of a colloid in a 32$^3$ system showing that the value, \texttt{colloid\_cell\_min} must be set such that it can contain the entire colloid plus colloid-colloid interactions.}
  \label{fig:colloid_int}
  \end{center}
\end{figure}

\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly, and as before, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}. There you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the files required for the initialisation and post-processing of colloids, respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.

\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}

\item In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
\item \texttt{cds\_with\_v  = 1} \\ which will output the positions and velocity data of the colloids,
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory.
\item For the post-processing of other data either copy the existing executable \texttt{extract} from 
previous examples or compile in \texttt{/util} in the usual manner.
Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
(see section \ref{configbuild}).

\item Copy the script \texttt{extract.py} as well into your working directory and set:
\begin{enumerate}
\item \texttt{vel = 1}, \\ to visualise the velocity field,
\item \texttt{q = 1}, \\ since there are liquid crystals present,
\item \texttt{colcds = 0} and 
\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
\end{enumerate}

\item Run the post-processing routine as before using the python script. 
\texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{lcs-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:

>>>>

\lstset{style=terminalverbatim}

@@

The following section outlines the configuration and compilation 
procedure. The configuration step is based on the GNU Compiler Collection gcc.
\begin{enumerate}
\item Go to the top level directory \texttt{/ludwig}:
\begin{lstlisting}
$ cd ludwig
\end{lstlisting}
\item Copy the (parallel) configuration file \texttt{lunix-mpicc-default.mk}
to top level directory 
and rename to \texttt{config.mk} (both steps exectued below at once):
\begin{lstlisting}
$ cp config/unix-mpicc-default.mk ./config.mk
\end{lstlisting}
Note that you may need to specify the correct path to the MPI wrapper compiler
in this file (assumed to be \texttt{mpicc}).

For production runs you may want to use thie compiler flag below instead of the default one:
\begin{lstlisting}
CFLAGS  = -O3 -g -DNDEBUG -DNSIMDVL=4
\end{lstlisting}
\item Issue \texttt{make} in the top level directory:
\begin{lstlisting}
$ make 
\end{lstlisting}
This creates the executable file \texttt{./Ludwig.exe} in \texttt{/src}. 
Note that more than one thread can be used in the compilation process,
which can lead to a considerable speedup  
\begin{lstlisting}
$ make -j 4
\end{lstlisting} 
If you want a serial build, clean your directory tree with
\begin{lstlisting}
$ make clean
\end{lstlisting}
use the conifguation file
\texttt{config/unix-gcc-default.mk} and include the extra keyword \texttt{serial}:
\begin{lstlisting}
$ make serial
\end{lstlisting}
\end{enumerate}

@@

In this section the Poiseuille flow of a simple fluid in a cavity, is outlined. 
The example is a step-by-step guide for performing a short run ($10,000$ time-steps) 
with \texttt{Ludwig} code, followed by a short introduction to the post-processing 
and data analysis procedures. Before starting, the code should be compiled in 
parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.
\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory \texttt{/docs/tutorial/test1} 
to your working directory. Note that the input files in this tutorial are modified versions of the 
reference input file \texttt{input.ref} in \texttt{/src} with a limited set of parameters
for each run.
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 4 ./Ludwig.exe input
\end{lstlisting}
\item You should now see statistics being reported on your standard output and 
output files being created with names of the form \texttt{vel-XXXXXXXX.001-001}. 
These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
\item The first step to the post-processing is to get hold of the \texttt{extract} executable in \texttt{/util},
which is created during the main compilation of \texttt{Ludwig}, and copy it into your tutorial working directory. 
\item Post-process the raw data by issuing 
\begin{lstlisting}
$ ./extract -a -i meta_file_name data_file_name_stub
\end{lstlisting} 
replacing the filenames with those of the present case (i.e. \texttt{vel.001-001.meta} and e.g. \texttt{vel-00010000}).
The command line options '\texttt{-a}' and '\texttt{-i}' give output in ASCII format and a lattice index 
for easy visualisation in \texttt{gnuplot}.
\item You should now see files named \texttt{vel-XXXXXXXX}. The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.
\item To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
the spatial $x$-coordinate on the lattice with the command:
\begin{lstlisting}
gnuplot> p `vel-XXXXXXXX' u 1:5
\end{lstlisting} 
By plotting the data in each file, you should be able to see the velocity profile converging to the final profile as shown in Fig. \ref{fig:velocityProfile}.

In the directory \texttt{/docs/tutorial/test1} a reference file of the final 
output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
\end{enumerate} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\linewidth]{velProf.png}
  \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps.}
  \label{fig:velocityProfile}
  \end{center}
\end{figure}

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test2} 
to your working directory. This input file contains the parameters required for this run. 
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item As before, you should now see statistics being reported on your standard output. But 
in this case two sets of output files are being created, one containing the velocity and 
the other compositional order parameter ($\phi$) data at each lattice point in the simulation. 
In this example, these data are in binary format.
\item You are able to use the same exectuable \texttt{extract} from the previous 
example for the  post-processing with different arguments and the \texttt{-k} flag to produce
files in VTK-format:
\begin{lstlisting}
$ ./extract -k vel.001-001.meta vel-00010000.001-001
$ ./extract -k phi.001-001.meta phi-00010000.001-001
\end{lstlisting}
\item A bash script like this one below can be used for convenient post-processing of multiple 
raw data files and time steps (with appropriate start, increment and end in the \texttt{for} loop):
\begin{lstlisting}
#!/bin/bash
for i in `seq 1000 1000 10000`;
do
  tstep=$(printf "%08d" $i)
  ./extract -k vel.001-001.meta vel-${tstep}.001-001  
  ./extract -k phi.001-001.meta phi-${tstep}.001-001
done
\end{lstlisting}
\item You should now see files with the extension \texttt{.vtk} corresponding to the velocity 
and $\phi$ data ready to be visualised using ParaView.
\begin{enumerate}
\item In Paraview open the files containing the $\phi$ data.
\item Click on the group of files in the pipeline browser and apply a contour map using 
Filters/Common/Contour or in the tool bar:

@@

For this third example, the system being used is a that of colloidal particles suspended in 
a liquid crystalline fluid.
The example illustrates the post-processing procedure for a colloidal system in addition to 
demonstrating the ways to visualise liquid crystals in ParaView. 
As before, the code should be compiled in parallel and the executable 
\texttt{Ludwig.exe} should be copied into your working directory.
\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test3} to your working directory. We see in this input file that 
there are many more parameters 
than in the previous examples. Specifically, we note that \texttt{colloid\_cell\_min} 
must be set if colloids are present and must be at least $2r + \delta$ where $r$ is colloid 
radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.
 \begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\linewidth]{colloidInit.png}
  \caption{A diagrammatic representation of a colloid in a 32$^3$ system showing that the value, \texttt{colloid\_cell\_min} must be set such that it can contain the entire colloid plus colloid-colloid interactions.}
  \label{fig:colloid_int}
  \end{center}
\end{figure}
\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly, and as before, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}. There you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the files required for the initialisation and post-processing of colloids, respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\item Now issue
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.
\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item An \texttt{extract\_colloids} executable is produced in \texttt{/util} during the main compilation of \texttt{Ludwig}. 
For our problem we need to set potentially different parameters. In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII format,
\item \texttt{cds\_with\_v  = 1} \\ which will output the positions and velocity data of the colloids,
\item Now issue
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory.
\item For the post-processing of other data either copy the existing executable \texttt{extract} from 
previous examples or compile in \texttt{/util} in the usual manner.
\item Run the post-processing routines as before using different arguments:
\begin{lstlisting}
$ ./extract -k vel.001-001.meta vel-00010000.001-001
$ ./extract -k -s -d q.001-001.meta q-00010000.001-001
$ ./extract_colloids config.cds00010000 1 col-cdsvel-00010000.csv
\end{lstlisting}
The \texttt{-s} and \texttt{-d} flags produce scalar order parameter and director field output.
\texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item Again, a bash script like this one below can be used for convenient post-processing of multiple raw data files and time steps (with appropriate start, increment and end in the \texttt{for} loop):
\begin{lstlisting}
#!/bin/bash
for i in `seq 1000 1000 10000`;
do
  tstep=$(printf "%08d" $i)
  ./extract -k vel.001-001.meta vel-${tstep}.001-001  
  ./extract -k -s -d q.001-001.meta q-${tstep}.001-001
  ./extract_colloids config.cds${tstep} 1 col-cdsvel-${tstep}.csv
done
\end{lstlisting}
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{lcs-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:
